---
title: "PA1"
author: "Sanaz Dehghani"
date: "3/18/2021"
output: html_document
---
Loading the data
```{r echo=TRUE}
activity<-read.csv("activity.csv")
```

Calculating the total number of steps taken per day
```{r echo=TRUE}
activityPERday <- aggregate(steps ~ date, data = activity, FUN = sum)
```

Making histogram of the total number of steps taken each day
```{r echo=TRUE}
hist(activityPERday$steps, xlab = "Steps per Day", main = "Total steps per day")
```
Calculating and reporting the mean and median of the total number of steps taken per day
```{r echo=TRUE}
meansteps <- format(mean(activityPERday$steps),digits = 1)
mediansteps <-format(median(activityPERday$steps),digits = 1) 
```

## What is the average daily activity pattern?
Making a time series plot 
```{r echo=TRUE}
averagesteps <- aggregate(steps ~ interval, data = activity, FUN = mean)
plot(averagesteps$interval, averagesteps$steps, type = "l", xlab = "Intervals", ylab = "Average steps", main = "Average steps per interval")
```
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo=TRUE}
maxsteps <-max(averagesteps$steps)
maxinterval <- averagesteps$interval[which(averagesteps$steps == maxsteps)]
maxsteps <- round(maxsteps, digits = 2)
```
##Imputing missing values
Calculating and reporting the total number of missing values in the dataset
```{r echo=TRUE}
sum(is.na(activity))
```
Devising a strategy for filling in all of the missing values in the dataset
and median total number of steps taken per day
```{r echo=TRUE}
# calculate mean of steps per interval, we end up with a mean for all 288 intervals
MeanStepsPerInterval <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
# cut the 'activity' dataset in 2 parts (with and without NAs)
activity_NAs <- activity[is.na(activity$steps),]
activity_non_NAs <- activity[!is.na(activity$steps),]
#replace missing values in activity_NAs
activity_NAs$steps <- as.factor(activity_NAs$interval)
levels(activity_NAs$steps) <- MeanStepsPerInterval
#change the vector back as integer 
levels(activity_NAs$steps) <- round(as.numeric(levels(activity_NAs$steps)))
activity_NAs$steps <- as.integer(as.vector(activity_NAs$steps))
#merge/rbind the two datasets together
imputed_activity <- rbind(activity_NAs, activity_non_NAs)
```
Making a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day
```{r echo=TRUE}
#Plotting parameters to place previous histogram and new one next to each other
par(mfrow = c(1,2))
#Plot again the histogram from the first part of the assignment
activity_steps_day <- aggregate(steps ~ date, data = activity, FUN = sum, na.rm = TRUE)
hist(activity_steps_day$steps, xlab = "Steps per Day", main = "NAs REMOVED - Total steps/day")
#Plot new histogram, with imputed missing values
imp_activity_steps_day <- aggregate(steps ~ date, data = imputed_activity, FUN = sum, na.rm = TRUE)
hist(imp_activity_steps_day$steps, xlab = "Steps per Day", main = "NAs IMPUTED - Total steps/day")
```
##Are there differences in activity patterns between weekdays and weekends?
indicating whether a given date is a weekday or weekend day
```{r echo=TRUE}
#elseif function to categorize Saturday and Sunday as factor level "weekend", all the rest as "weekday"
imputed_activity$dayType <- ifelse(weekdays(as.Date(imputed_activity$date)) == "Samstag" | weekdays(as.Date(imputed_activity$date)) == "Sonntag", "weekend", "weekday")
#transform dayType variable into factor
imputed_activity$dayType <- factor(imputed_activity$dayType)
#Aggregate a table showing mean steps for all intervals, acrlss week days and weekend days
steps_interval_dayType <- aggregate(steps ~ interval + dayType, data = imputed_activity, FUN = mean)
#verify new dataframe 
head(steps_interval_dayType)
```
making a panel plot containing a time series plot 
```{r echo=TRUE}
#add descriptive variables
names(steps_interval_dayType) <- c("interval", "day_type", "mean_steps")
#plot with ggplot2
library(ggplot2)
plot <- ggplot(steps_interval_dayType, aes(interval, mean_steps))
plot + geom_line(color = "tan2") + facet_grid(day_type~.) + labs(x = "Intervals", y = "Average Steps", title = "Activity Patterns")
```

